Conversations in time: interactive visualisation to explore structured temporal data

### 2020-09-24 tsibbletalk.Rmd
slicing

The shared tsibble data leverages the `key` attribute to converse with many coordinated views, with or without \CRANpkg{shiny}. On the other hand, a second critical attribute--`index`--lays the foundational temporal context that augments the conversation. When temporal data are plotted and stretched against the entire span like Figure \ref{fig:highlight-retail-1}, it puts emphasis on the trend perception. Yet to digest periodic/aperiodic patterns, data should be wrapped over relative time units that are origin-less, such as one quarter or one day.

```{r wrap-ped, fig.show = "hold", fig.cap = "The shiny module for wrapping the \\code{pedestrian20} data.", fig.subcap = c("Initial overview state", "1-day state", "7-day state, anchoring to Monday"), fig.ncol = 1}
include_graphics("img/wrap-0.png")
include_graphics("img/wrap-1.png")
include_graphics("img/wrap-7.png")
```

The city of Melbourne has sensors installed to count hourly tallies of pedestrians in order to capture downtown daily rhythms [@ped]. Figure \ref{fig:wrap-ped} shows the first five months of 2020 foot traffic at four locations, with the depiction of three pronounced slices in time. Figure \ref{fig:wrap-ped-1} unfolds all counts from January to May on their absolute time line, facetted by four sensors. On March 16, Melbourne went to the stage three lockdown due to COVID-19, seeing a significant decline in traffic volume across the city. These lines are then folded into daily and weekly sections, shown as Figure \ref{fig:wrap-ped-2} and \ref{fig:wrap-ped-3} respectively. Seasonal variations have been popped out to viewers, complementing the not-just-magnitude-drop story. The pre-lockdown period is coloured with dark green and lockdown with orange.

The wrapping procedure involves slicing time indices into seasonal periods of interest and their corresponding time dices. For example, hourly pedestrian data can be decomposed into 24-hour blocks grouped by all respective days, like Figure \ref{fig:wrap-ped-2}. Figure \ref{fig:wrap-ped} suggests that there could be more than one eye-catching slices out of many possible combinations, and thus repeated wrappings can be unwieldy. To visually locate a salient slice, the \CRANpkg{tsibbletalk} package implements a shiny module, a pair of UI and server functions, to automate this wrapping procedure.

This shiny module, decoupled to `tsibbleWrapUI()` and `tsibbleWrapServer()`, presents a clean interface and forms a resusable piece in a shiny application. A shiny module provides the vehicle in modularising shiny applications for both users and developers. Like all shiny modules, the first argument in both functions requires a user-supplied session id that must be unique. The UI function `tsibbleWrapUI()` simply shows a slider that animates or controls the number of periods to be diced. The workhorse is certainly the server function `tsibbleWrapServer()`, encapsulating the algorithm that transforms data and sends messages to update the plot accordingly. The `plot` argument expects a `ggplot` or `plotly` object, where one can plot data using either lines or other graphical elements (such as boxplots). As the function name suggests, a (shared) tsibble is needed to start the engine, and thereby the time `index` can be retrieved for dissection. The `period` option semantically takes a desired number of seasonal periods to be shifted, for example data shifted by "1 day", "2 days", or "1 week", etc. In other words, the `period` defines the grind level. For date-times (represented by `POSIXt`), the granularity ranges from fine "day" to coarse "year". The following code snippet generates Figure \ref{fig:wrap-ped}.

@@

Upon running the shiny application, Figure \ref{fig:wrap-ped-1} corresponds to the initial state, with the slider incremented by 1-day unit. The "play" button near the end of slider begins animating the slicing and dicing process by walking through all 24 hours by 152 days. Alternatively, users can drag the handler to poke around certain slices themselves.

In response to the slider input, the plot will be updated and loaded with newly transformed data. At its core, keeping the application as performant as possible is the top priority. Without completely redrawing the plot, the `plotlyProxy()` react method is invoked internally for talking to shiny. The underlying tsibble data is being called back and processed in R. Only transformed data gets fed back to the shiny server, along with reseting the x-axis ranges and breaks. The rest plot configurations, such as marks, y-axes, and layouts, are properly cached.

The new shiny module exploits the temporal aspect for a tsibble object, exposed by the `index` attribute. It allows users to slide through relative periods to digest seasonal behaviours, with slick user experience.

>>>>

An important aspect of temporal data is the time context. Time has cyclical structure, that may correspond to seasonal patterns to be discovered. The `index` component of the shared tsibble data forms the basis for exploring seasonality. To investigate for periodic or aperiodic patterns, series should be wrapped on themselves, where the index is broken into temporal components like quarter or day. We'll explore this with pedestrian traffic in Melbourne, Australia.

<!-- The shared tsibble data leverages the `key` attribute to converse with many coordinated views, with or without \CRANpkg{shiny}. On the other hand, a second critical attribute--`index`--lays the foundational temporal context that augments the conversation. When temporal data are plotted and stretched against the entire span like Figure \ref{fig:highlight-retail-1}, it puts emphasis on the trend perception. Yet to digest periodic/aperiodic patterns, data should be wrapped over relative time units that are origin-less, such as one quarter or one day.-->

```{r wrap-ped, fig.show = "hold", fig.cap = "Snapshots wrapping after slicing the \\code{pedestrian20} data at different intervals, none (top), daily (middle) and weekly (bottom). This type of interaction is made possible with Shiny elements.", fig.subcap = c("Initial overview state", "1-day state", "7-day state, anchoring to Monday"), fig.ncol = 1}
include_graphics("img/wrap-0.png")
include_graphics("img/wrap-1.png")
include_graphics("img/wrap-7.png")
```

The city of Melbourne has sensors installed at various locations, to record hourly counts of pedestrians, in order to capture daily rhythms of the downtown [@ped]. Figure \ref{fig:wrap-ped} shows the first five months of 2020 foot traffic at four different locations, for three different time slices, daily, weekly and full five months. Plot \ref{fig:wrap-ped-1} shows hourly counts from January to May on an absolute time line, facetted by location. The stage 3 COVID-19 lockdown, on March 16, is marked by a change of colour. (The pre-lockdown period is coloured with dark green and lockdown with orange.) We can see a significant decline in foot traffic  at all four locations. QV Market is less affected probably because this is a major produce market, an essential service that continued to operate. Bourke St, a major shopping center, sees a gradual uptick in the last weeks of the period indicating that people were getting back into the shops.

Figure \ref{fig:wrap-ped-2} and \ref{fig:wrap-ped-3} show slicing and wrapping of the series into daily and weekly sections, respectively. Multiple seasonality pops out. There tends to be a daily pattern, especially visible at the main train station, Southern Cross. There is also a week day vs weekend pattern, also most visible at Southern Cross. These seasonal patterns are still present during the lockdown, but the magnitude is greatly reduced. Numbers are also down at the produce market and the shopping center. Birrarung Marr is the most affected. This is the location of special events and its clear that these have completely disappeared during lockdown.

The wrapping procedure involves slicing the time index into seasonal periods of interest, and the result is diced time. For example, hourly pedestrian data can be decomposed into 24-hour blocks, which then overlays the counts for all respective days, as done in plot \ref{fig:wrap-ped-2}. For exploration, this slice position should be controlled interactively, so that many different slices can be examined rapidly. This can be achieved using shiny, with the functions provided in a \CRANpkg{tsibbletalk}.

This shiny module, decoupled to `tsibbleWrapUI()` and `tsibbleWrapServer()`, presents a clean interface and forms a reusable code that could be embedded in any shiny application. In general, a shiny module provides a vehicle for modularising shiny applications, relevant for both users and developers. As with all shiny modules, the first argument in both functions in \CRANpkg{tsibbletalk} requires a user-supplied session id that must be unique. The UI function `tsibbleWrapUI()` simply shows a slider that animates or controls the number of periods to be diced. The workhorse is the server function `tsibbleWrapServer()`, encapsulating the algorithm that transforms data and sends messages to update the plot accordingly. The `plot` argument expects a `ggplot` or `plotly` object, where one can plot data using either lines or other graphical elements (such as boxplots). As the function name suggests, a (shared) tsibble is needed to start the engine, so that the time `index` can be retrieved for dissection. The `period` option semantically takes a desired number of seasonal periods to be shifted, for example data shifted by "1 day", "2 days", or "1 week", etc. In other words, the `period` defines the grind level. For date-times (represented by `POSIXt`), the granularity ranges from fine "day" to a much coarser "year". The following code snippet generates Figure \ref{fig:wrap-ped}.

@@

Figure \ref{fig:wrap-ped-1} corresponds to the initial state, with the slider incremented by 1-day units. The "play" button near the end of slider can automately animate the slicing and dicing process, walking the viewer through all 24 hours of the 152 days. Alternatively, users can drag the slider to examine selected slices.

In response to the slider input, the plot will be updated and loaded with newly transformed data. At its core, keeping the application as performant as possible is the top priority. Without completely redrawing the plot, the `plotlyProxy()` react method is invoked internally for talking to shiny. The underlying tsibble data is being called back and processed in R. Only transformed data gets fed back to the shiny server, for updating with resetting the x-axis ranges and breaks. The other plot configurations, such as marks, y-axes, and layouts, are cached and used as is.

The new shiny module exploits the temporal aspect for a tsibble object, available throu the `index` attribute. It allows users to slide through relative periods to digest seasonal behaviours, with nimble user experience.

### 2020-09-24 tsibbletalk.tex
slicing

The shared tsibble data leverages the \texttt{key} attribute to converse
with many coordinated views, with or without \CRANpkg{shiny}. On the
other hand, a second critical attribute--\texttt{index}--lays the
foundational temporal context that augments the conversation. When
temporal data are plotted and stretched against the entire span like
Figure \ref{fig:highlight-retail-1}, it puts emphasis on the trend
perception. Yet to digest periodic/aperiodic patterns, data should be
wrapped over relative time units that are origin-less, such as one
quarter or one day.

@@

\caption[The shiny module for wrapping the \code{pedestrian20} data]{The shiny module for wrapping the \code{pedestrian20} data.}\label{fig:wrap-ped}
\end{figure}
\end{Schunk}

The city of Melbourne has sensors installed to count hourly tallies of
pedestrians in order to capture downtown daily rhythms \citep{ped}.
Figure \ref{fig:wrap-ped} shows the first five months of 2020 foot
traffic at four locations, with the depiction of three pronounced slices
in time. Figure \ref{fig:wrap-ped-1} unfolds all counts from January to
May on their absolute time line, facetted by four sensors. On March 16,
Melbourne went to the stage three lockdown due to COVID-19, seeing a
significant decline in traffic volume across the city. These lines are
then folded into daily and weekly sections, shown as Figure
\ref{fig:wrap-ped-2} and \ref{fig:wrap-ped-3} respectively. Seasonal
variations have been popped out to viewers, complementing the
not-just-magnitude-drop story. The pre-lockdown period is coloured with
dark green and lockdown with orange.

The wrapping procedure involves slicing time indices into seasonal
periods of interest and their corresponding time dices. For example,
hourly pedestrian data can be decomposed into 24-hour blocks grouped by
all respective days, like Figure \ref{fig:wrap-ped-2}. Figure
\ref{fig:wrap-ped} suggests that there could be more than one
eye-catching slices out of many possible combinations, and thus repeated
wrappings can be unwieldy. To visually locate a salient slice, the
\CRANpkg{tsibbletalk} package implements a shiny module, a pair of UI
and server functions, to automate this wrapping procedure.

This shiny module, decoupled to \texttt{tsibbleWrapUI()} and
\texttt{tsibbleWrapServer()}, presents a clean interface and forms a
resusable piece in a shiny application. A shiny module provides the
vehicle in modularising shiny applications for both users and
developers. Like all shiny modules, the first argument in both functions
requires a user-supplied session id that must be unique. The UI function
\texttt{tsibbleWrapUI()} simply shows a slider that animates or controls
the number of periods to be diced. The workhorse is certainly the server
function \texttt{tsibbleWrapServer()}, encapsulating the algorithm that
transforms data and sends messages to update the plot accordingly. The
\texttt{plot} argument expects a \texttt{ggplot} or \texttt{plotly}
object, where one can plot data using either lines or other graphical
elements (such as boxplots). As the function name suggests, a (shared)
tsibble is needed to start the engine, and thereby the time
\texttt{index} can be retrieved for dissection. The \texttt{period}
option semantically takes a desired number of seasonal periods to be
shifted, for example data shifted by ``1 day'', ``2 days'', or ``1
week'', etc. In other words, the \texttt{period} defines the grind
level. For date-times (represented by \texttt{POSIXt}), the granularity
ranges from fine ``day'' to coarse ``year''. The following code snippet
generates Figure \ref{fig:wrap-ped}.

@@

Upon running the shiny application, Figure \ref{fig:wrap-ped-1}
corresponds to the initial state, with the slider incremented by 1-day
unit. The ``play'' button near the end of slider begins animating the
slicing and dicing process by walking through all 24 hours by 152 days.
Alternatively, users can drag the handler to poke around certain slices
themselves.

In response to the slider input, the plot will be updated and loaded
with newly transformed data. At its core, keeping the application as
performant as possible is the top priority. Without completely redrawing
the plot, the \texttt{plotlyProxy()} react method is invoked internally
for talking to shiny. The underlying tsibble data is being called back
and processed in R. Only transformed data gets fed back to the shiny
server, along with reseting the x-axis ranges and breaks. The rest plot
configurations, such as marks, y-axes, and layouts, are properly cached.

The new shiny module exploits the temporal aspect for a tsibble object,
exposed by the \texttt{index} attribute. It allows users to slide
through relative periods to digest seasonal behaviours, with slick user
experience.

>>>>

An important aspect of temporal data is the time context. Time has
cyclical structure, that may correspond to seasonal patterns to be
discovered. The \texttt{index} component of the shared tsibble data
forms the basis for exploring seasonality. To investigate for periodic
or aperiodic patterns, series should be wrapped on themselves, where the
index is broken into temporal components like quarter or day. We'll
explore this with pedestrian traffic in Melbourne, Australia.

@@

\caption[Snapshots wrapping after slicing the \code{pedestrian20} data at different intervals, none (top), daily (middle) and weekly (bottom)]{Snapshots wrapping after slicing the \code{pedestrian20} data at different intervals, none (top), daily (middle) and weekly (bottom). This type of interaction is made possible with Shiny elements.}\label{fig:wrap-ped}
\end{figure}
\end{Schunk}

The city of Melbourne has sensors installed at various locations, to
record hourly counts of pedestrians, in order to capture daily rhythms
of the downtown \citep{ped}. Figure \ref{fig:wrap-ped} shows the first
five months of 2020 foot traffic at four different locations, for three
different time slices, daily, weekly and full five months. Plot
\ref{fig:wrap-ped-1} shows hourly counts from January to May on an
absolute time line, facetted by location. The stage 3 COVID-19 lockdown,
on March 16, is marked by a change of colour. (The pre-lockdown period
is coloured with dark green and lockdown with orange.) We can see a
significant decline in foot traffic at all four locations. QV Market is
less affected probably because this is a major produce market, an
essential service that continued to operate. Bourke St, a major shopping
center, sees a gradual uptick in the last weeks of the period indicating
that people were getting back into the shops.

Figure \ref{fig:wrap-ped-2} and \ref{fig:wrap-ped-3} show slicing and
wrapping of the series into daily and weekly sections, respectively.
Multiple seasonality pops out. There tends to be a daily pattern,
especially visible at the main train station, Southern Cross. There is
also a week day vs weekend pattern, also most visible at Southern Cross.
These seasonal patterns are still present during the lockdown, but the
magnitude is greatly reduced. Numbers are also down at the produce
market and the shopping center. Birrarung Marr is the most affected.
This is the location of special events and its clear that these have
completely disappeared during lockdown.

The wrapping procedure involves slicing the time index into seasonal
periods of interest, and the result is diced time. For example, hourly
pedestrian data can be decomposed into 24-hour blocks, which then
overlays the counts for all respective days, as done in plot
\ref{fig:wrap-ped-2}. For exploration, this slice position should be
controlled interactively, so that many different slices can be examined
rapidly. This can be achieved using shiny, with the functions provided
in a \CRANpkg{tsibbletalk}.

This shiny module, decoupled to \texttt{tsibbleWrapUI()} and
\texttt{tsibbleWrapServer()}, presents a clean interface and forms a
reusable code that could be embedded in any shiny application. In
general, a shiny module provides a vehicle for modularising shiny
applications, relevant for both users and developers. As with all shiny
modules, the first argument in both functions in \CRANpkg{tsibbletalk}
requires a user-supplied session id that must be unique. The UI function
\texttt{tsibbleWrapUI()} simply shows a slider that animates or controls
the number of periods to be diced. The workhorse is the server function
\texttt{tsibbleWrapServer()}, encapsulating the algorithm that
transforms data and sends messages to update the plot accordingly. The
\texttt{plot} argument expects a \texttt{ggplot} or \texttt{plotly}
object, where one can plot data using either lines or other graphical
elements (such as boxplots). As the function name suggests, a (shared)
tsibble is needed to start the engine, so that the time \texttt{index}
can be retrieved for dissection. The \texttt{period} option semantically
takes a desired number of seasonal periods to be shifted, for example
data shifted by ``1 day'', ``2 days'', or ``1 week'', etc. In other
words, the \texttt{period} defines the grind level. For date-times
(represented by \texttt{POSIXt}), the granularity ranges from fine
``day'' to a much coarser ``year''. The following code snippet generates
Figure \ref{fig:wrap-ped}.

@@

Figure \ref{fig:wrap-ped-1} corresponds to the initial state, with the
slider incremented by 1-day units. The ``play'' button near the end of
slider can automately animate the slicing and dicing process, walking
the viewer through all 24 hours of the 152 days. Alternatively, users
can drag the slider to examine selected slices.

In response to the slider input, the plot will be updated and loaded
with newly transformed data. At its core, keeping the application as
performant as possible is the top priority. Without completely redrawing
the plot, the \texttt{plotlyProxy()} react method is invoked internally
for talking to shiny. The underlying tsibble data is being called back
and processed in R. Only transformed data gets fed back to the shiny
server, for updating with resetting the x-axis ranges and breaks. The
other plot configurations, such as marks, y-axes, and layouts, are
cached and used as is.

The new shiny module exploits the temporal aspect for a tsibble object,
available throu the \texttt{index} attribute. It allows users to slide
through relative periods to digest seasonal behaviours, with nimble user
experience.
